Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-09 latex/main.tex
Updates from Overleaf

We find strong evidence for the recombination events leading to 17 Pango Xs in
the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango Xs are XA, which
is the first designated Pango X that showed signs of community transmission
\cite{Jackson2021}, and XBB, which escaped vaccine-induced immunity and gained
global prevalence \cite{Tamura2023}. The recombinant origin of XA is well
explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A;
Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and
breakpoint location.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
recombination node (A); the recombination node associated with a group of
``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

\subsection*{Quantification of evidence for recombination}
%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021} (Table \ref{tab:jackson}).

@@

\subsection*{Recombination breakpoint intervals along the genome}
We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Methods).

@@

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{recombinant_intervals}
    \caption{Recombination intervals. (A) Distribution of the breakpoint
intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top). Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
    \label{fig:recombinant_intervals}
\end{figure}

@@

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}
The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples, building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article with Python code alone and
no compiled languages.

@@

\subsubsection*{Comparing phylogenetic backbones}

@@

\subsubsection*{Mutational spectra}
To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024}, and
a global UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are relative proportions of each type of nucleotide mutation.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs (e.g., B.1.1.7-like and B.1.1.7-like+E484K for Alpha).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes which have a Scorpio label associated with Alpha, Delta, BA.1, BA.2, BA.4, or BA.5
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using Nextclade and has samples collected globally up to November, 2022.
We used the following mapping between the WHO labels and the Nextstrain labels:
Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
The mutation count data are available at https://github.com/jbloomlab/SARS2-mut-spectrum/.

@@

\subsubsection*{Concordance among recombination detection methods}
For RecombinHunt, we obtained two sets of detection results. 
One set is based on mutational profiles calculated from a global GISAID dataset, 
and the other set from a global Nextstrain dataset. 
We adjusted the RecombinHunt-identified breakpoint interval 
so that they are right-exclusive by adding one to the right coordinate.

For CovRecomb, we summarized the detection results for the Pango Xs as follows. 
First, we obtained the "independent" recombination events 
corresponding to the Pango Xs (if available) 
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
When multiple events corresponded to a Pango X, 
we took the results of the event having the most epidemic recombinants. 
Then, for other Pango Xs, we collated the data from a file (URL) 
containing all the detection results produced by the study. 
For each of these Pango Xs, 
we took the most commonly observed combination of 
left parent, right parent, and breakpoint interval.
We adjusted the CovRecomb-identified breakpoint interval 
so that they are right-exclusive by adding one to the right coordinate.

We ran rebar on the sampled sequences associated with each robust recombination event. 
Rebar identifies a sequence as a recombinant if it contains sets of mutations 
predetermined to be associated with different Pango lineages.

The parents of a Pango X inferred using a method are considered concordant 
if their Pango lineage assignments are at least as specific as the community-identified parents.
Also, a breakpoint interval of a Pango X inferred using a method is considered concordant 
if it overlaps with the community-identified breakpoint interval by at least one base.
We adjusted the breakpoint intervals from the community 
by subtract one from the left coordinate and adding two to the right coordinate.

\subsubsection*{Calculating Pango lineage distance using pangonet}
We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
The idea is to count the number of edges separating two given Pango lineages in a ``pseudo-phylogeny''
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages in the ``alias\_key.json''
and ``lineage\_notes.txt'' files
from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

\subsection{Analysis of the Jackson et al. recombinants}
Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha, followed
by an analysis to reduce the list of plausible recombinants displaying signs of
onward transmission. Here we compared our sc2ts detection results of these
recombinants with the detection results reported by Jackson et al.

@@

\begin{figure}
    \centering
    %TODO fix this to make the figures smaller.
    \includegraphics[width=\linewidth]{mutational_spectra}
    \caption{All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
    the mutation count data from Bloom et al. (2023) (B), and
    the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
    \label{fig:mutational_spectra}
}
\end{figure}

>>>>

We find strong evidence for the recombination events leading to 17 Pango Xs in
the ARG (Table~\ref{tab:pango_x_lineages}), all of which pass the 
robustness filtering criteria (Figure~\ref{fig:recombinant_qc}). 
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango Xs are XA,
which is the first designated Pango X that showed signs of onward transmission\cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence\cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
recombination node (A); the recombination node associated with a group of
``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

\subsection*{Quantification of evidence for recombination}

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021} (Table \ref{tab:jackson}).

@@

\subsection*{Recombination breakpoint intervals along the genome}

We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Methods).

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{recombinant_intervals}
    \caption{Recombination intervals. (A) Distribution of the breakpoint
intervals of the 356 robust recombination events along the genome
(annotated with gene labels on top). Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
    \label{fig:recombinant_intervals}
\end{figure}

@@

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods\cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples, building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article with Python code alone and
no compiled languages.

@@

\subsubsection*{Comparison of phylogenetic backbones}

@@

\subsubsection*{Comparison of mutational spectra}
To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are defined as the relative proportions of each possible type of nucleotide mutation.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering all the mutations above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs
(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes that have a Scorpio label associated with these VOCs
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using Nextclade and has samples collected globally up to November, 2022.
We used the following mapping between the WHO labels and the Nextstrain clade labels:
Alpha, 20I; Detal, 21I and 21J; BA.1, 21K; BA.2, 21L; BA.4, 22A, and BA.5, 22B.
The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.

@@

\subsubsection*{Comparison with recombination detection methods}
For RecombinHunt, we obtained two sets of detection results of the Pango X lineages
from the supplementary materials of Alfonso et al.\cite{Alfonsi2024}.
One set of results is based on mutational profiles calculated from a global GISAID dataset,
and the other set from a global NextStrain dataset.
The breakpoint intervals identified using RecombinHunt are defined
as the left and right recombination-informative sites.
We adjusted their coordinates by adding one to the right coordinate
so that the intervals are right-exclusive.

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
which correspond to the Pango X lineages (if available)
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango Xs,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

We ran rebar on the sampled sequences associated with each robust recombination event. 
Rebar identifies a sequence as a recombinant if it contains sets of mutations 
which are pre-determined to be associated with different Pango lineages.

The parents of a Pango X inferred using a method are deemed concordant
if they have Pango lineage assignments
that are at least as specific as those proposed by the community.
The breakpoint interval of a Pango X inferred using a method is deemed concordant
if it overlaps with that proposed by the community by at least one base.
These breakpoint intervals are defined as ranges of possible site positions
where a breakpoint is inferred to occur.
To make these breakpoint intervals right-exclusive,
we adjusted them by subtracting one from the left coordinate and adding two to the right coordinate.

\subsubsection*{Calculation of Pango lineage distance using pangonet}
We define a ``pseudo-phylogenetic'' distance to quantify the divergence between two Pango lineages.
The idea is to count the number of edges separating two given Pango lineages in a ``pseudo-phylogeny''
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages in the ``alias\_key.json''
and ``lineage\_notes.txt'' files
from the Pango designation issues (https://github.com/cov-lineages/pango-designation).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

\subsection{Analysis of the Jackson et al. recombinants}

Jackson et al.\cite{Jackson2021} conducted detailed analyses of early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha,
followed by an analysis to reduce the list of plausible recombinants
displaying signs of onward transmission.
Here we compared the sc2ts detection results of these recombinants
with the detection results reported by Jackson et al.

@@

\begin{figure}
    \centering
    %TODO fix this to make the figures smaller.
    \includegraphics[width=\linewidth]{mutational_spectra}
    \caption{
    All-site mutational spectra of major VOCs calculated from the sc2ts ARG (A),
    the mutation count data from Bloom et al. (2023)\cite{Bloom2023} (B), and
    the Viridian UShER phylogeny from Hunt et al. (2024)\cite{Hunt2024}.
    \label{fig:mutational_spectra}
}
\end{figure}
